Using GIT with GitHub
======================

Today we will be working with one of the most popular version control systems called [Git][id_git].
Like [R][id_r], [Git][id_git] is free and open source, decreasing the boundaries for global collaborations.
We'll be operating [Git][id_git] from Terminal and using [GitHub][id_github] as our remote online repository.

### GIT works in four main layers:
```{r echo=FALSE, message=FALSE, warning=FALSE, comment='', dpi=200, out.width='504px', out.height='504px'}
plot.info  <-  function(y, text1, text2, text3, final.arrows) {
    polygon(c(1,5,5,1), c(y,y,y+1,y+1), col='grey90')
    text(3,y+0.3,text1,pos=3)
    arrows(8,y+0.5,6,y+0.5)
    text(8,y+0.5,text2,pos=4)
    if(final.arrows) {
        arrows(3,y-0.1,3,y-0.9, col='red')
        text(3.5,y-0.5,text3,pos=4,cex=1.1,col='red')
    }
}

chart.info  <-  list(c(9,'working directory','just a folder on your computer','$ git add',TRUE), 
                     c(7,'stage','get files ready to commit','$ git commit',TRUE),
                     c(5,'local repository','the permanent history of changes','$ git push',TRUE),
                     c(3,'remote repository','online repository (shared?)','',FALSE))

par(family='Monaco', mar=c(0,0,0,0), cex=0.9)
plot(0, 0, type='n', xlim=c(0,15), ylim=c(3,10),  ylab='', xlab='', axes=FALSE)
x<-sapply(chart.info, function(x){plot.info(as.numeric(x[1]),x[2],x[3],x[4],x[5])})

```

### Step 1: create [GitHub][id_github] account
* using your academic account has advantages - free private repositories

### Step 2: set up global user configurations on Terminal
* `git config --global user.name "your desired username"`
* `git config --global user.email "your email (academic has advantages)"`

### Step 3: create a new local repo in shell dir:
```{r eval=FALSE} 
$ git init
```

### Step 4: check status of repo:
```{r eval=FALSE} 
$ git status
```

### Step 5: create a new file README.md and add it (stage) to the local repo:
```{r eval=FALSE}
$ git add filename

#conversely you can add all files (and deletions) by typing
$ git add --all
```

### Step 6: recheck status

### Step 7: ready to commit the change? 
```{r eval=FALSE}
$ git commit -m 'your message here'
```

### Step 8: check commit's history:
```{r eval=FALSE} 
$ git log
```

### Step 9: we want you to gather in pairs, one student will be *student A*, and the other *student B*. Let's collaborate with [GitHub][id_github]:
* student A:
    * create repo on [GitHub][id_github]
    * copy and paste last two lines of code from [GitHub][id_github] to add remote/push
    * look at repo page on [GitHub][id_github]
    * add person B as a collaborator (settings > collaborators)
* student B:
    * clone person A's repo
        * at the bottom right side of the repo page, click on 'You can clone with HTTPS' and then copy the link shown right above 
    * on your machine, open the repo folder, find the README.md file, open in RStudio and make a change to it
    * open Terminal and stage the file by typing `$ git add`
    * commit the change with the message `$ git commit -m 'changed README'`
    * push to the repo on GitHub typing `$ git push`
* student A:
    * pull the latest changes made by student B by typing `$ git pull`

[id_git]: http://git-scm.com/
[id_r]: http://www.r-project.org
[id_github]: https://github.com/