Control Flow in R
==================
    
#### Often when we're coding we want to control the flow of our actions. That can be done by setting things to happen only if a condition or a set of conditions are met. Alternatively, we can also set an action to be computed for a particular number of times. There are many ways these can be achieved in R. For conditional statements, the most commonly used approaches are the constructs: 
```{r eval=FALSE} 
if (and else)
while
```

#### Say, for example, that we want R to print a message depending on how a number x relates to 5:
```{r}
x  <-  5
if(x >= 5) {
    print(paste0(x, ' is greater or equal than 5'))
} else {
    print(paste0(x, 'is smaller than 5'))
}
```

#### We could also have a situation where we want an action to be executed while a condition is being matched:
```{r}
x   <-  0
vec <-  vector()
while(x < 10) {
    vec <- c(vec, x^2)
    x   <- x+1
}
vec
```

#### Now, and this is very important. When R evaluates the condition inside if and while statements, it is looking for a logical element, i.e., TRUE or FALSE. For example:
```{r}
if(4 == 3){
    '4 equals 3'
}
```

#### The message was not printed because the condition 4 == 3 is FALSE 
```{r}
4==3
```

#### R also offers a built-in function that is more intuitive to evaluate conditions.
```{r}
mat  <-  matrix(0:19,5,4)
mat  <-  ifelse(mat < 5, 0, mat)

#OR

mat  <-  matrix(sample(c(NA, 1:10),50,replace=TRUE), 5, 10)
mat  <-  ifelse(is.na(mat), 0, 1)
```

#### In the first example, ifelse looks for all the conditions there are TRUE in the statement `mat < 5` and replace them by the condition, otherwise does the altervative action. In this case, I told R to keep the original `mat` values where the condition did not occur. Similarly, in the second example, I told R to replace all existing numbers with 1 and replace missing values with 0. *VERY IMPORTANT*: ifelse can be very handy, but for heavy and repetitive tasks, it is considerably slower than regular `if` and `else` statements. 

#### As above, another way of controlling the flow of actions is to set them to occur for a specific number of times. We achieve that using the command `for`
```{r}
salad    <-  c('lettuce', 'carrots', 'tomatos')
amounts  <-  c('1 bunch', '1 package', '1 bag')
for (vegetable in seq_along(salad)) {
    print(paste0('Our salad has ', amounts[vegetable], ' of ', salad[vegetable]))
}
```

#### Notice that at each loop, the index `vegetable` takes a value following the sequence of the vector salad. During the first loop, vegetable, in this particular case, is going to be 1, then 2, and finally 3. 
```{r}
for (index in 1:3) {
    print(index)
}
```

#### The index can take any name you want (preferably names that do not overlap with existing objects or functions). Conventionally, when we deal with matrices or data.frames, we use the index `i` to denote rows and the index `j` to denote columns. Example:
```{r}
mat  <-  matrix(1:4,2,2)
mat
for (i in 1:nrow(mat)) {
    for(j in 1:ncol(mat)) {
        print(mat[i,j])
    }
}
```

#### At this point you may be wondering if we can mix the different approaches and the answer is yes. For a certain `for` loop, you may want to compute your action only if a condition is matched.
```{r}
set.seed(10)
mat  <-  matrix(rpois(16,lambda=11), 4, 4)
mat
for (i in 1:nrow(mat)) {
    for(j in 1:ncol(mat)) {
        if(mat[i,j] >= 10) {
            next
        } else {
            print(paste0('element at row ', i, ', column ', j, ', is smaller than 5'))
        }
    }
}
```

#### Do not worry that much about the the functions `set.seed` and `rpois`. All you need to know is that `set.seed` will allow us to generate the same random numbers in different computers for comparison matters and `rpois` generates positive integers randomly following a Poisson distribution. Notice that I used `next` to skip the loop in case the condition was TRUE. You can also use the command `break` if you want to stop the loop at once. 

## Exercise
### Step 1: create a matrix called `mat.ex` with 3 rows and 10 columns using `set.seed(20); rpois(30, lambda=12)` as values;
### Step 2: create an empty numeric vector called `means` with length = ncol(mat.ex)
### Step 3: loop through the columns of `mat.ex` and calculate the respective mean values of each column;
### Step 4: if the mean calculated value is greater than 11, store it in `means`;
### Step 5: print vector `means` without zeros

#### solution:
```{r}
set.seed(20)
mat.ex  <-  matrix(rpois(30, lambda=12), 3, 10)
means   <-  vector(mode='numeric', length=ncol(mat.ex))
for(j in 1:ncol(mat.ex)) {
    if(mean(mat.ex[,j]) > 11)
        means[j] <- mean(mat.ex[,j])
}
print(means[means != 0])
```

# Speeding up using the family apply
#### One of the common features across most R beginners is to think everything as a loop. While this approach is valid and in many cases the fastest solution to think of, sometimes it can be very, very slow. Try the following:
```{r}
set.seed(10)
x  <-  matrix(rpois(500, lambda=25), 20, 25)
a  <-  vector(mode='numeric',length=length(x))
for(index in 1:ncol(x)) {
    a[index] <- mean(x[,index])
}
```

#### That took < 1 second to run. Although it was effectively fast, the code was relatively big (5 lines) for its small task. Fortunately, R has a family of loop functions called the `apply` family, which are slightly different but generally as fast as or faster loops than `for` loops. For the moment, you do not need to fully understand the detailed differences between a regular `for` loop and it's equivalent using the `apply` family. Let's replicate the example above using the mother function `apply`:
```{r}
set.seed(10)
x  <-  matrix(rpois(500, lambda=25), 20, 25)
a  <-  apply(x, 2, mean)
```

#### `a` takes the same values as in the example before, but with much less code (3 lines) which is also simpler. In this case, `apply` evaluates the object `x` and then apply the function `mean` to its 2nd dimension, i.e., columns - dimension=1 would imply rows. `apply` assumes no margin for dimension for it's evaluations, so you always have to provide it. There are even more convinient and faster ways to calculate sum or mean of dimensions in a matrix or data.frame.
```{r}
set.seed(10)
x  <-  matrix(rpois(500, lambda=25), 20, 25)
a  <-  colMeans(x) #or colSums or rowMeans or rowSums
```

#### Finally, you can customize your own functions and tomorrow John is going to teach you how to embed your customized functions inside `apply` calls. 
