Control Flow in R
==================
    
#### Often when we're coding we want to control the flow of our actions. Control flow is simply the order in which we code and have our statements evaluated. That can be done by setting things to happen only if a condition or a set of conditions are met. Alternatively, we can also set an action to be computed for a particular number of times. There are many ways these can be achieved in R. For conditional statements, the most commonly used approaches are the constructs: 
```{r eval=FALSE} 
if (and else)
while
```

#### Say, for example, that we want R to print a message depending on how a number x relates to 5:
```{r}
x  <-  5
if(x >= 5) {
    print(paste0(x, ' is greater or equal than 5'))
} else {
    print(paste0(x, 'is smaller than 5'))
}
```

#### We could also have a situation where we want an action to be executed while a condition is being matched:
```{r}
x   <-  0
vec <-  vector()
while(x < 10) {
    vec <- c(vec, x^2)
    x   <- x+1
}
vec
```

#### Now, and this is very important. When R evaluates the condition inside if and while statements, it is looking for a logical element, i.e., TRUE or FALSE. For example:
```{r}
if(4 == 3){
    '4 equals 3'
}
```

#### The message was not printed because the condition 4 == 3 is FALSE 
```{r}
4==3
```

#### R also offers a built-in function that is more intuitive to evaluate conditions.
```{r}
mat  <-  matrix(0:19,5,4)
mat  <-  ifelse(mat < 5, 0, mat)

#OR

mat  <-  matrix(sample(c(NA, 1:10),50,replace=TRUE), 5, 10)
mat  <-  ifelse(is.na(mat), 0, 1)
```

#### In the first example, ifelse looks for all the conditions there are TRUE in the statement `mat < 5` and replaces them by the first action (transform in 0), otherwise does the altervative action (keep originals). In this case, I told R to keep the original `mat` values where the condition did not occur. Similarly, in the second example, I told R to replace all existing numbers with 1 and replace missing values with 0. *VERY IMPORTANT*: ifelse can be very handy, but for heavy and repetitive tasks, it is considerably slower than regular `if` and `else` statements. 

#### As above, another way of controlling the flow of actions is to set them to occur for a specific number of times. We achieve that using the command `for`
```{r}
salad    <-  c('lettuce', 'carrots', 'tomatos')
amounts  <-  c('1 bunch', '1 package', '1 bag')
for (vegetable in seq_along(salad)) {
    print(paste0('Our salad has ', amounts[vegetable], ' of ', salad[vegetable]))
}
```

#### Notice that at each loop, the index `vegetable` takes a value following the sequence of the vector salad. During the first loop, vegetable, in this particular case, is going to be 1, then 2, and finally 3. 
```{r}
for (index in 1:3) {
    print(index)
}
```

#### The index can take any name you want (preferably names that do not overlap with existing objects or functions). Conventionally, when we deal with matrices or data.frames, we use the index `i` to denote rows and the index `j` to denote columns. Example:
```{r}
mat  <-  matrix(1:4,2,2)
mat
for (i in 1:nrow(mat)) {
    for(j in 1:ncol(mat)) {
        print(mat[i,j])
    }
}
```

#### At this point you may be wondering if we can mix the different approaches and the answer is yes. For a certain `for` loop, you may want to compute your action only if a condition is matched.
```{r}
set.seed(10)
mat  <-  matrix(rpois(16,lambda=11), 4, 4)
mat
for (i in 1:nrow(mat)) {
    for(j in 1:ncol(mat)) {
        if(mat[i,j] >= 10) {
            next
        } else {
            print(paste0('element at row ', i, ', column ', j, ', is smaller than 5'))
        }
    }
}
```

#### Do not worry that much about the the functions `set.seed` and `rpois`. All you need to know is that `set.seed` will allow us to generate the same random numbers in different computers for comparison matters and `rpois` generates positive integers randomly following a Poisson distribution. Notice that I used `next` to skip the loop in case the condition was TRUE. You can also use the command `break` if you want to stop the loop at once. 

## Exercise
### Step 1: create a matrix called `mat.ex` with 3 rows and 10 columns using `set.seed(20); rpois(30, lambda=12)` as values;
### Step 2: create an empty numeric vector called `means` with length = ncol(mat.ex)
### Step 3: loop through the columns of `mat.ex` and calculate the respective mean values of each column;
### Step 4: if the mean calculated value is greater than 11, store it in `means`;
### Step 5: print vector `means` without zeros

#### solution:
```{r}
set.seed(20)
mat.ex  <-  matrix(rpois(30, lambda=12), 3, 10)
means   <-  vector(mode='numeric', length=ncol(mat.ex))
for(j in 1:ncol(mat.ex)) {
    if(mean(mat.ex[,j]) > 11)
        means[j] <- mean(mat.ex[,j])
}
print(means[means != 0])
```

# Optmizing tasks using the family apply
#### Suppose we have read in a data frame that contains gene expression samples of 5 genes for 2 species:
```{r}
set.seed(10)
expression.data  <-  data.frame(species=rep(c("Homosap",'Musmusc'), each=15), genes=rep(rep(paste('gene_',1:3), each=5), 2), norm.expression=rnorm(30, mean=15, sd=5), stringsAsFactors=FALSE)
expression.data
```

#### Now imagine that you want to extract meaninful numbers out of that table. Say, you want to analyse the mean and standard deviation of each gene per species:
```{r}
mean.sp  <-  tapply(expression.data$norm.expression, list(expression.data$species, expression.data$genes), mean)
sd.sp  <-  tapply(expression.data$norm.expression, list(expression.data$species, expression.data$genes), sd)
boxplot(expression.data$norm.expression ~ expression.data$species + expression.data$genes)
```

#### We can also use the `apply` function to calculate rows or column means directly from a matrix `x`, without having to do a loop.
```{r}
set.seed(10)
x   <-  matrix(rpois(500, lambda=25), 20, 25)
a1  <-  vector()
for(j in 1:ncol(x)) {
    a1 <- c(a1, mean(x[,j]))
}

#we avoid this loop by using apply:
a2  <-  apply(x, 2, mean) #or colSums or rowMeans or rowSums
#or even better (and faster!)
a3  <-  colMeans(x) #or colSums or rowMeans or rowSums
```

#### Finally, you can customize your own functions and tomorrow John is going to teach you how to embed your customized functions inside `apply` calls. 

# Dealing with lists
#### R offers a very powerful class of objects called lists. A list can virtually contain anything. You can have many data.frames inside a list, or you can mix a vector, a data.frame, any other type of array, other lists, and so on. Example:
```{r}
set.seed(10)
our_list  <-  list(runif(10), matrix(1:10,2,5), data.frame(letters=letters, order=1:26))
class(our_list)
str(our_list)
dim(our_list) #lists do not have dimensions as arrays, thay are measured by their number of elements
length(our_list)
names(our_list) #create names for our list
names(our_list)  <-  c('random_numbers', 'a_matrix', 'a_dataframe')
our_list
```

#### If we want to extract an element from a list, we use double squared brackets [[]]. 

```{r}
our_list[[1]]
our_list[['a_matrix']]
class(our_list[[1]])
dim(our_list[[2]])
```

#### Now that we understand the basics of a list, we can use use the function `lapply` from the `apply` family, which is a very powerful tool to loop over vectors and lists. Exmaple:
```{r}
list_gcs <- list(gene_a=runif(3), gene_b=runif(3), gene_c=runif(3))
means    <-  lapply(list_gcs, mean)
class(means)
names(means)
means[[1]] #or means[['gene_a']], for example
means
```

#### Using `our_list`, we can also use lapply to quickly explore the classes of our different objects within it.
```{r}
lapply(our_list, class)
```

### Notice that lapply returns a list as its default behavior. R offers a second function that simplifies the output whenever possible. It's called `sapply`, which stands for simplify `lapply`. If your input is a mtrix, the output will also be coerced into a matrix, if your input is a vector, R will try to coerce the output into a vector as well.
```{r}
sap  <-  sapply(our_list, class)
sap
class(sap)
```